A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-20 tidyindex.qmd
writing up the confidence interval example

## Every distribution, every scale, every index all at once

@@

## Confidence interval

@@

@fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia. Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2). early warning for the bushfire?

also the ABS one - could display some uncertainties

```{r fig-conf-interval}
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI) for the Texas post office station, in Queensland, Australia. A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and calculate the index. Since the start of 2019, both confidence intervals indicate an extreme drought condition (below -2)."
res <- queensland %>% 
  filter(name == "TEXAS POST OFFICE") %>% 
  init(id = id, time = ym) %>% 
  aggregate(.var = prcp, .scale = 24) %>% 
  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
  augment(.var = .agg)

res_band <- res$data %>% 
  group_by(ym, id, name) %>% 
  summarise(
    q025 = quantile(.index, 0.025),
    q05 = quantile(.index, 0.05),
    q10 = quantile(.index, 0.1),
    q50 = quantile(.index, 0.5),
    q90 = quantile(.index, 0.9),
    q95 = quantile(.index, 0.95),
    q975 = quantile(.index, 0.975)
    ) %>% 
  ungroup()

@@

The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework. The pipeline offers a modular workflow to allow compute index with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties. The framework proposed in the paper is universal to index across diverse domains. Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions and to illustrate how slight adjustment of linear combination weights impact the index.

>>>>

## Every distribution, every scale, every index all at once {#sec-example1}

@@

## Confidence interval [TO BE CHANGED]

Index uncertainties may arise from various stages of the pipeline, including from different parameterisation choices, as illustrated from @sec-example1, or from the statistical procedures in the pipeline. Given the lack of direct instruments for measurement, true values for indexes can never be obtained to compare the index against. Nevertheless, an index can be provided with confidence interval to report it accuracy. In this example, the Texas post office station highlighted in @fig-compute-temporal is used to illustrate the calculation of confidence interval for the Standardized Precipitation Index (SPI). Bootstrapping is used to account for the sampling uncertainty in the distribution fit step of the index pipeline and to assess its impact on the SPI series.

In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month. This results in  31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set of distribution parameters. To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series. In the `tidyindex` package, this bootstrap sampling is activated when the argument `.n_boot` is set to a value other than the default of 1. In the following code, the Standardized Precipitation Index (SPI) is calculated using a time scale of 24. The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations ( `.n_boot = 100`) and then fits the gamma distribution. The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.

@@

The confidence interval can then be calculated using the quantile method from the bootstrap samples. @fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia. From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought. The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought/ flood severity in extreme events.

```{r fig-conf-interval}
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia. A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and to calculate the index. The dashed line at SPI = -2 represents an extreme drought as defined by the SPI. Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line (-2) and are relatively wide compared to other time periods. This suggests that while it is certain that the Texas post office is suffering from a drastic drought, there is considerable uncertainty in quantifying its severity, given the extremity of the event."
res <- queensland %>% 
  filter(name == "TEXAS POST OFFICE") %>% 
  init(id = id, time = ym) %>% 
  aggregate(.var = prcp, .scale = 24) %>% 
  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
  augment(.var = .agg)

res_band <- res$data %>% 
  group_by(ym, id, name) %>% 
  summarise(
    q025 = quantile(.index, 0.025),
    q10 = quantile(.index, 0.1),
    q90 = quantile(.index, 0.9),
    q975 = quantile(.index, 0.975)
    ) %>% 
  ungroup()

@@

The paper introduces a data pipeline comprising nine modules designed for the construction and analysis of indexes within the tidy framework. The pipeline offers a modular workflow to allow compute index with different parameterizations, to test minor changes to the original index definition, and to quantify uncertainties. The framework proposed in the paper is universal to index across diverse domains. Examples are provided, including the drought indexes (SPI and SPEI) and Global Gender Gap Index (GGGI), to demonstrate the index calculation with different time scales and distributions, to illustrate how slight adjustment of linear combination weights impact the index, and to calculate confidence intervals on the index.

<!-- When adopting the pipeline approach to construct indexes, analysts may consider developing software that can be readily deployed in the cloud for production purposes. -->

### 2023-08-20 tidyindex.tex
writing up the confidence interval example

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{every-distribution-every-scale-every-index-all-at-once}{%
\subsection{Every distribution, every scale, every index all at
once}\label{every-distribution-every-scale-every-index-all-at-once}}

@@

\hypertarget{confidence-interval}{%
\subsection{Confidence interval}\label{confidence-interval}}

@@

Figure~\ref{fig-conf-interval} presents the 80\% and 95\% confidence
interval for the Texas post office station, in Queensland, Australia.
Since the start of 2019, both confidence intervals indicate an extreme
drought condition (below -2). early warning for the bushfire?

also the ABS one - could display some uncertainties

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
the Standardized Precipitation Index (SPI) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated pricipitation series to estimate gamma parameters
and calculate the index. Since the start of 2019, both confidence
intervals indicate an extreme drought condition (below -2).}

@@

The paper introduces a data pipeline comprising nine modules designed
for the construction and analysis of indexes within the tidy framework.
The pipeline offers a modular workflow to allow compute index with
different parameterizations, to test minor changes to the original index
definition, and to quantify uncertainties. The framework proposed in the
paper is universal to index across diverse domains. Examples are
provided, including the drought indexes (SPI and SPEI) and Global Gender
Gap Index (GGGI), to demonstrate the index calculation with different
time scales and distributions and to illustrate how slight adjustment of
linear combination weights impact the index.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{sec-example1}{%
\subsection{Every distribution, every scale, every index all at
once}\label{sec-example1}}

@@

\hypertarget{confidence-interval-to-be-changed}{%
\subsection{Confidence interval {[}TO BE
CHANGED{]}}\label{confidence-interval-to-be-changed}}

Index uncertainties may arise from various stages of the pipeline,
including from different parameterisation choices, as illustrated from
Section~\ref{sec-example1}, or from the statistical procedures in the
pipeline. Given the lack of direct instruments for measurement, true
values for indexes can never be obtained to compare the index against.
Nevertheless, an index can be provided with confidence interval to
report it accuracy. In this example, the Texas post office station
highlighted in Figure~\ref{fig-compute-temporal} is used to illustrate
the calculation of confidence interval for the Standardized
Precipitation Index (SPI). Bootstrapping is used to account for the
sampling uncertainty in the distribution fit step of the index pipeline
and to assess its impact on the SPI series.

In SPI, the distribution fit step fits the gamma distribution to the
aggregated precipitation series separately for each month. This results
in 31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set
of distribution parameters. To account for this sampling uncertainty
with these samples, bootstrapping is used to generate replicates of the
aggregated series. In the \texttt{tidyindex} package, this bootstrap
sampling is activated when the argument \texttt{.n\_boot} is set to a
value other than the default of 1. In the following code, the
Standardized Precipitation Index (SPI) is calculated using a time scale
of 24. The bootstrap procedure samples the aggregated precipitation
(\texttt{.agg}) for 100 iterations ( \texttt{.n\_boot\ =\ 100}) and then
fits the gamma distribution. The resulting gamma probabilities are then
transformed into normal densities in the normalizing step with
\texttt{agument()}.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. The relatively wide confidence
interval, as well as during the excessive precipitation events in
1996-1998 and 1999-2000, suggests a high variation of the gamma
parameters estimated from the bootstrap samples and its difficulty to
accurately quantifying the drought/ flood severity in extreme events.

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
the Standardized Precipitation Index (SPI-24) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated pricipitation series to estimate gamma parameters
and to calculate the index. The dashed line at SPI = -2 represents an
extreme drought as defined by the SPI. Most parts of the confidence
intervals from 2019 to 2020 sit below the extreme drought line (-2) and
are relatively wide compared to other time periods. This suggests that
while it is certain that the Texas post office is suffering from a
drastic drought, there is considerable uncertainty in quantifying its
severity, given the extremity of the event.}

@@

The paper introduces a data pipeline comprising nine modules designed
for the construction and analysis of indexes within the tidy framework.
The pipeline offers a modular workflow to allow compute index with
different parameterizations, to test minor changes to the original index
definition, and to quantify uncertainties. The framework proposed in the
paper is universal to index across diverse domains. Examples are
provided, including the drought indexes (SPI and SPEI) and Global Gender
Gap Index (GGGI), to demonstrate the index calculation with different
time scales and distributions, to illustrate how slight adjustment of
linear combination weights impact the index, and to calculate confidence
intervals on the index.
